---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First, we will load the data and preview it.
```{r echo=TRUE}
options("scipen" = 10)
unzip('activity.zip')
data <- read.csv('activity.csv')
str(data[1:5,])
```
Next, we will transform the data variable into the appropriate format.
```{r  echo=TRUE}
data$date <- as.Date(data$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

```{r echo=TRUE}

# Calculate the mean no. of steps per day
totalsteps <- aggregate(steps ~ date, data = data, sum)

# Calculate median across the means
mediansteps <- round(median(totalsteps$steps),0)

# Calculate the mean across the means
meansteps <- round(mean(totalsteps$steps),0)


# Plot histogram and overlay the mean/median
library(ggplot2)
ggplot(totalsteps, aes(x=steps)) +
  geom_histogram(binwidth = 1000, color = "darkblue", fill = "lightblue") +
  geom_vline(xintercept = meansteps, col = "black") +
  geom_text(aes(x=meansteps, label = "Mean/Median", y = 8), angle = 90, vjust = 1.2) +
  ggtitle("Figure 1: Distribution of total number of steps taken each day") +
  labs(x = "Total number of steps taken per day", y = "Frequency")

dev.copy(png, './figure/plot1.png', width = 480, height = 480)
dev.off()
``` 
The mean number of steps per day is `r meansteps` and the median number of steps per day is `r mediansteps`.

## What is the average daily activity pattern?
```{r echo=TRUE}
# Calculate the mean number of steps for each interval
avg <- aggregate(steps ~ interval, data = data, FUN = mean)

# Identify the maximum point
maxpoint <- avg[which.max(avg$steps),]
  
# Plot the time series
ggplot(avg, aes(x=interval, y=steps)) +
  geom_line() +
  geom_point(data = maxpoint, aes(x=interval, y=steps), color = "red", size = 3) +
  geom_text(data = maxpoint, aes(x=interval, y = steps, label = paste0("Maximum (",interval,",",round(steps,0),")"))) +
  ggtitle("Figure 2: Average daily activity across time intervals") +
  labs(x = "Interval", y = "Average daily activity (no. of steps)")

dev.copy(png, './figure/plot2.png', width = 480, height = 480)
dev.off()
```
The maximum number of steps is contained within the `r maxpoint$interval` interval.

## Imputing missing values

We will inspect the number of NAs for each variable.
```{r echo=TRUE}
na <- with(data, sapply(data, function(x) {mean(is.na(x))}))
na
```
About 13% of the data are missing values. We will proceed to impute them by taking the mean for that 5-minute interval, which we had earlier computed.

```{r echo=TRUE}
library(dplyr)

# Replace missing data with mean of that interval
imputeddata <- data %>%
                mutate(steps = ifelse(is.na(steps), avg[avg$interval == interval,]$steps, steps))
```

Now we update the histogram with the imputed data.
```{r echo=TRUE}

# Re-calculate the mean number of steps per day
totalsteps2 <- aggregate(steps ~ date, data = imputeddata, sum)

# Re-calculate the median of the means
mediansteps2 <- round(median(totalsteps2$steps),0)

# Re-calculate the mean of the means
meansteps2 <- round(mean(totalsteps2$steps),0)

# Plot histogram and overlay with the mean/median
ggplot(totalsteps2, aes(x=steps)) +
  geom_histogram(binwidth = 1000, color = "darkblue", fill = "lightblue") +
  geom_vline(xintercept = meansteps2, col = "black") +
  geom_text(aes(x=meansteps2, label = "Mean/Median", y = 8), angle = 90, vjust = 1.2) +
  ggtitle("Figure 3: Distribution of total number of steps taken each day") +
  labs(x = "Total number of steps taken per day", y = "Frequency")

dev.copy(png, './figure/plot3.png', width = 480, height = 480)
dev.off()
```
The updated mean number of steps per day is `r meansteps2` and the median number of steps per day is `r mediansteps2`, which remains generally unchanged as our imputation method applies the mean value and therefore preserves the overall mean.

## Are there differences in activity patterns between weekdays and weekends?

```{r echo=TRUE}

# Create factor variable indicating whether the day is a weekday or weekend
imputeddata$dayofweek <- weekdays(imputeddata$date)
imputeddata$weekdayindicator <- as.factor(ifelse(imputeddata$dayofweek %in% c("Saturday", "Sunday"), "weekend", "weekday"))

# Calculate average steps by interval and weekday/weekend
avgbydayofweek <- aggregate(steps ~ interval + weekdayindicator, data = imputeddata, mean)

# Plot time series
ggplot(avgbydayofweek, aes(x=interval, y=steps)) +
  geom_line() +
  facet_grid(weekdayindicator~.) +
  ggtitle("Figure 4: Average daily activity across time intervals") +
  labs(x = "Interval", y = "Average daily activity (no. of steps)")

dev.copy(png, './figure/plot4.png', width = 480, height = 480)
dev.off()
```
There is a higher peak in activity during the weekdays than weekends. However, during the weekends, activity seems to be generally higher across the day.